Fast Sorting on Modern Computers

### 2005-08-18 docs/tech_report/heap.tex
done spell check. some errors found
merged davids final changes in
reread predictor section and added comparison between H_N and actual results for short arrays

\section{A More Detailed Look at Branch Prediction}

@@

Figure \ref{base and memory-tuned heapsorts-1}(a) shows the behaviour of each
branch in the base heapsort algorithm. The `\n{Creation, Has Children}' branch
refers to the check as to whether a key has children, during the creation of the
heap The `\n{Creation, Child Comparison}' branch is the comparison between the
two children in the 2-heap, and the `\n{Creation, Child Promotion}' refers to
checking if the child is larger than it's parent. The same checks are later made
during the destruction of the heap, and are referred to as `\n{Destruction, Has
Children}', `\n{Destruction, Child Comparison}' and `\n{Destruction Child
Promotion}'.

In the memory-tuned heapsort, the creation phase is simpler, involving merely
one branch, called `\n{Fix-up}' here. The other columns refer to the
destruction phase, where most of the work takes place in memory-tuned heapsoirt.
When a 4-heap is now used, there are three comparisons instead of just one.
Likewise, in an (8-heap), there are seven comparisons.

%TODO read over the merged bits. might not be quite right

The results of the software predictor show the relation of the branches
in the creation and destruction of the heap. The base heapsort results in
Figure \ref{base and memory-tuned heapsorts-1} show that the time taken to
create the heap is not significant relative to the time taken to sort from the
heap.

@@

More importantly, the accuracy of the each successive comparison increases in
the same way as selection sort (see Section \ref{selection branches harmonics}),
becoming more predictable as the fanout increases. For example, in the 8-heap,
shown in Figures \ref{base and memory-tuned heapsorts-3} and \ref{base and
memory-tuned heapsorts-4}, the first comparison is predicted correctly 50\% of
the time, while the next is predicted correctly 61\% of the time, then 71\%,
77\%, and so on in a curve, which is shown in Figure \ref{memory-tuned heapsort
curve}.

@@

\label{heapsort entropy}
We observe here that the more branches there are, the more are correctly
predicted, and that there is a threshold below which the number of branches will
not drop. This is discussed in more Section \ref{quicksort entropy}.

Also notable from the results is that branches are slightly more often not-taken
then they are correctly predicted, implying that a semi-static predictor based
on hints from the compiler could do a better job at making correct predictions.

@@

>>>>

\section{A More Detailed Look at Branch Prediction}
\label{heapsort-branch-prediction}

@@

Figure \ref{base and memory-tuned heapsorts-1} shows the behaviour of each
branch in the base heapsort algorithm. The `\n{Creation, Has Children}' branch
refers to the check as to whether a key has children, during the creation of the
heap. The `\n{Creation, Child Comparison}' branch is the comparison between the
two children in the 2-heap, and the `\n{Creation, Child Promotion}' refers to
checking if the child is larger than it's parent. The same checks are later made
during the destruction of the heap, and are referred to as `\n{Destruction, Has
Children}', `\n{Destruction, Child Comparison}' and `\n{Destruction, Child
Promotion}'.

In the memory-tuned heapsort, the creation phase is simpler, involving merely
one branch, called `\n{Fix-up}' here. The other columns refer to the destruction
phase, where most of the work takes place in memory-tuned heapsort.  When a
4-heap is used (see Figure \ref{base and memory-tuned heapsorts-2}), there are
three comparisons instead of just one. Similarly, in an 8-heap (Figures
\ref{base and memory-tuned heapsorts-3} and \ref{base and memory-tuned
heapsorts-4}), there are seven comparisons.

The results of the software predictor show the relation of the branches
in the creation and destruction of the heap. The base heapsort results show that
the time taken to create the heap is not significant relative to the time taken
to sort from the heap.

@@

More importantly, the accuracy of the each successive comparison increases in
the same way as selection sort (see Section \ref{selection branches harmonics}),
becoming more predictable as the fanout increases. For example, in the 8-heap,
the first comparison is predicted correctly 50\% of the time, while the next is
predicted correctly 61\% of the time, then 71\%, 78\%, and so on in a curve,
which is shown in Figure \ref{memory-tuned heapsort curve}.

The relationship between this number and $H_N-1$ is relatively straightforward.
$H_N-1$ represents the cumulative number of changes in the minimum. Over a long
array, this approximates the number of branch mispredictions, as almost every
branch will not be taken, and the predictor will always predict not-taken. The
graph shows us the rate of correct predictions, and each point is calculated
using a seperate branch predictor. As a result, we expect $1-y = \frac{1}{x+1}$,
where $x$ and $y$ are the cartesian coordinates of each point on the graph.
$1-y$ is the number of mispredictions from the graph, while $\frac{1}{x+1}$ is
the increase of $H_N$ for step $N$.

When the array is short, as in this example, the branch predictor has too much
feedback to be as accurate as $H_N$. For $N=2$, we would approximate a 33\%
misprediction ratio using $H_N$, but we measure 39\%. The approximated and
actual ratios become closer as $N$ increases, as we would expect, and for $N=7$,
we approximate 12.5\% and measure 12.7\%.

@@

\label{heapsort entropy}
We observe also that the more branches there are, the more are correctly
predicted, and that there may a threshold below which the number of branches will
not drop. This is discussed further in Section \ref{quicksort entropy}.

@@

It should be possible to more accurately model and therefore predict the branch
prediction ratio when the array is short. Our results show us what we expect the
answer to be, but have not generated a formula to predict it.
